Hydras, Ordinals \& Co. \\ A library in Coq of entertaining formal mathematics

### 2021-09-27 doc/JFLA2022/paper.tex
improve (?) Section 2

\begin{abstract}
  \emph{Hydra-battles} is a collaborative library of commented proofs, developped under the \emph{Coq-Community} project.
  We present its evolution since a former presentation~\cite{JFLA2018paper} and how its structure is determined by other
  libraries and the \emph{Alectryon} literate proving tool.
\end{abstract}

@@

We also study the number of steps of a battle:
Let $\alpha<\epsilon_0$ be an ordinal.
Using the \texttt{coq-equations} plug-in~\cite{sozeau:hal-01671777}, we define the function which returns the number of steps of the battle starting with $\alpha$ at step $i$, for  any $i$ in $\mathbb{N}$.
We prove that this number is greater or equal than
$H'_\alpha(i)-i$, where $H'$ is a slight variant of the Hardy hierarchy of rapidly growing functions~\cite{BW85, KS81, Promel2013, Wainer1970}, defined by transfinite recursion in Fig.~\ref{fig:Hprime}.
From this result, we infer that the considered function is not primitive recursive in general (for $\alpha\geq\omega^\omega$).

Using $H'$s equations as rewrite rules, we can study a concrete example of hydra battle. Let us start with the hydra of figure~\ref{fig:start} and $0$ as initial step.
By a combination of rewriting and induction, we prove that the number of steps of the considered battle has at least $2^{10^{21}}$ binary digits.

@@

\begin{figure}[h]
 \noindent
 For any $\alpha$ in $[0,\epsilon_0)$ and any $i$ in $\mathbb{N}$,
   \begin{itemize}
\item If $\alpha=0$, then $H'_\alpha (i)= i$.
\item If $\alpha$ is a limit ordinal, then 
  $H'_\alpha(i) = H'_{(\canonseq{\alpha}{i+1})}(i)$.
  \item If $\alpha=\beta+1$, then 
$H'_\alpha(i)=H'_\beta(i+1)$.
\end{itemize}
\vspace{4pt}
\inputsnippets{HprimeDef}
\caption{The rapidly growing hierarchy $H'$ of arithmetic functions}
\label{fig:Hprime}
\end{figure}

@@

@@

Computability is a key topic in Computer Science teaching. Moreover O'Connor's library is a nice illustration of dependently typed programming, so we chose to devote a full chapter to this formalisation, with comments on the definitions and proofs, [counter]-examples and exercises.

>>>>

\begin{abstract}
  \emph{Hydra-battles} is a collaborative library of commented proofs, developped under the \emph{Coq-Community} project.
  We present its evolution since a former presentation and how its structure is determined by other
  libraries and the \emph{Alectryon} literate proving tool.
\end{abstract}

@@

We also study the number of steps of a battle:
Let $\alpha<\epsilon_0$ be an ordinal. 
We prove that  the number of steps of the battle starting with
$\alpha$ at step $i$ is greater or equal than
$H'_\alpha(i)-i$, where $H'$ is a slight variant of the Hardy hierarchy of rapidly growing functions~\cite{BW85, KS81, Promel2013, Wainer1970}.  The function $H'_\alpha$ is defined by transfinite recursion over $\alpha$:

\begin{align}
  H'_0(i) & = i\\
  H'_\alpha(i) &= H'_{(\canonseq{\alpha}{i+1})}(i)  \quad\textit{if $\alpha$ is a limit ordinal}\\
  H'_{\alpha}(i) &=H'_\beta(i+1) \quad\textit{if $\alpha=\beta+1$}
\end{align}

Figure~\ref{fig:Hprime} 
 shows the  definition of $H'$ with the \texttt{coq-equations}
 plug-in~\cite{sozeau:hal-01671777}.

 \begin{figure}[h]
 \inputsnippets{HprimeDef}
\caption{Coq Equations for $H'$}
\label{fig:Hprime}
\end{figure}

Using $H'$s equations as rewrite rules, we can study a realistic example. We take the hydra of figure~\ref{fig:start} and $i=0$ as initial configuration. 
By a sequence of rewritings and inductions, we prove that the number of steps of the considered battle is greater or equal than $2^{2^N}$, where $N=2^{70}-1$.

@@

@@

More generally, we prove that, for $\alpha\geq\omega^\omega$
the function computing the length of the battle starting with the configuration $(\alpha,i)$ is not primitive recursive.

@@

Computability is a key topic in Computer Science teaching. Moreover O'Connor's library is a nice illustration of dependently typed programming, so we chose to devote a full chapter to this formalization, with comments on the definitions and proofs, [counter]-examples and exercises.
